---
title: "Heart Rate Variability and Psychosocial Stress"
subtitle: "An Atherosclerotic Risk in Communities Study"
author: 
- Anish Shah, MD^[Department of Medicine, School of Medicine, Emory University]
- Amit Shah, MD, MSCR^[Department of Epidemiology, Rollins School of Public Health, Emory University]
data: "`r format(Sys.time(), '%B %d, %Y')`"
fontsize: 10pt
bibliography: hrv.bib
csl: american-medical-association.csl
link-citations: yes
output: 
  beamer_presentation:
    colortheme: "beaver"
    slide_level: 3
  latex_engine: xelatex
header-includes:
  - \usepackage{dcolumn}
  - \usepackage{float}
editor_options: 
  chunk_output_type: console
---

```{r setup, global_options, include=FALSE}
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(
  cache = TRUE,
  warning = FALSE,
  eval = TRUE,
  echo = FALSE,
  include = TRUE,
  message = FALSE,
  dpi = 600,
  dev = "png",
  options("scipen" = 0, "digits" = 3),
  tinytex.verbose = TRUE,
  tidy = FALSE
)

options(xtable.comment = FALSE)
```

```{r, include=FALSE}

# Libraries
source("1_libraries.R")

# Data intake
source("2_intake.R")

# Tidy
source("3_tidy.R")
```

# Overview

### Disclosures and Funding

- Emory University
- Atherosclerosis Risk in Communities Study
- No conflicts of interest

### Atherosclerosis Risk in Communities Study

The ARIC study is a prospective epidemiologic study designed to investigate atherosclerosis and its clinical outcomes by cardiovascular risk factors, diseases, and demographics. It involved longitudinal follow-up over five visits from 1987 to 2013, with outcomes updated as of 2015, with over 15k initial participants. It collected measures of psychosocial stress (increased anger, increased vital exhaustion, decreased social support) and HRV over time.

### Background

Psychosocial stressors, such as fatigue and vital exhaustion, are repeatedly shown to be associated independently with MACE.[@Appels1988; @Bogle2018] Stressors and depression are shown to correlate with lower HRV.[@Huang2018; @Vroege2012; @Shah2013] The relationship between autonomic dysfunction and psychosocial stressors is still being studied. 

### Purpose

- Identify relationship of HRV at V1 to psychosocial states at V2
- Identify relationship of psychosocial states at V2 with HRV at V4 (cross-lagged analysis)
- Control for influence of gender on HRV measures
- Identify longitudinal relationship of anger on HRV

# Description of data 

## HRV considerations

### Heart rate

Heart rate and variability interact, more so mathematically than clinically, and such, can be done mathematically before they are interpreted.[@Sacha2014a] 

- geometric domains are divided by RR in seconds
- frequency domains are divided by RR^2 in seconds

### Gender

Gender differences in HRV exist, with differences in their prognostic/clinical value.[@Sacha2014a] Its been suggested that in women, LF/HF is lower, HF is higher, LF is lower, VLF is lower, and SDNN is lower.[@Koenig2016] We believe these two groups should be analyzed separately as the mechanism is still unknown.

```{r, results='asis', include=FALSE}
# Table for V1
df <- inner_join(hrv1, cs1, by = "ID") %>% na.omit()
tabular( (n=1) + Format(digits = 2)*(LF_HF + HFc + LFc + VLFc + TPc + SDNNc + RMSSDc)*(mean + sd) ~ (GENDER), data = df )
tmp_v1 <- 
  df %>%
  dplyr::select(ID, HFc, LFc, VLFc, TPc, LF_HF, SDNNc, RMSSDc, PNN50, GENDER) %>%
  gather(key = hrv, value = value, -ID, -GENDER) %>%
  group_by(GENDER, hrv) %>%
  summarise(value = list(value)) %>%
  spread(GENDER, value) %>%
  group_by(hrv) %>%
  mutate(p.value = t.test(unlist(Female), unlist(Male))$p.value,
         t.value = t.test(unlist(Female), unlist(Male))$statistic)

# Table for V4
df <- inner_join(hrv4, cs4, by = "ID") %>% inner_join(., cs1[c("ID", "GENDER")]) %>% na.omit()
tabular( (n=1) + Format(digits = 2)*(LF_HF + HFc + LFc + VLFc + TPc + SDNNc + RMSSDc)*(mean + sd) ~ (GENDER), data = df )
tmp_v4 <- 
  df %>%
  dplyr::select(ID, HFc, LFc, VLFc, TPc, LF_HF, SDNNc, RMSSDc, PNN50, GENDER) %>%
  gather(key = hrv, value = value, -ID, -GENDER) %>%
  group_by(GENDER, hrv) %>%
  summarise(value = list(value)) %>%
  spread(GENDER, value) %>%
  group_by(hrv) %>%
  mutate(p.value = t.test(unlist(Female), unlist(Male))$p.value,
         t.value = t.test(unlist(Female), unlist(Male))$statistic)
```

After correction for HR, we found that in both V1 (n=`r length(cs1$ID)`) and V4 (n=`r length(cs4$ID)`), the following differences in similar magnitude:

- HF (p=`r tmp_v4$p.value[tmp_v4$hrv == "HFc"]`]) contrary to prior studies[@Koenig2016]
- LF (p=$`r tmp_v4$p.value[tmp_v4$hrv == "LFc"]`$) 
- VLF (p=$`r tmp_v4$p.value[tmp_v4$hrv == "VLFc"]`$) ECG were 2m at V1, 5m at V4
- LF/HF (p=$`r tmp_v4$p.value[tmp_v4$hrv == "LF_HF"]`$) 
- SDNN (p=$`r tmp_v4$p.value[tmp_v4$hrv == "SDNNc"]`$)
- PNN50 (p=$`r tmp_v4$p.value[tmp_v4$hrv == "PNN50c"]`$)

### HRV measures at V1

```{r}
# For males and females
tmp <- bind_rows(z_hrv1_male, z_hrv1_female)

# Create SEM
df <-
  cs1[c("ID", "HTN", "CHD", "DIABETES", "GENDER")] %>%
  gather(key = covar, value = status, -ID, -GENDER) %>%
  inner_join(gather(data = tmp[-3], key = hrv, value = value, -ID), ., by = "ID") %>%
  group_by(GENDER, covar, status, hrv) %>%
  na.omit() %>%
  summarise(mean = mean(value), n = length(value), sd = sd(value), se = sd(value) / sqrt(length(value)))

# Visualize
ggplot(df, aes(x=hrv, colour=hrv, shape = status)) + 
  facet_wrap(GENDER ~ covar) +  
  geom_point(aes(y = mean), size=3, position=position_dodge(0.3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2, position=position_dodge(0.3)) +
  theme_minimal() +
  scale_colour_viridis(option = "cividis", discrete = TRUE, begin = 0.1, end = 0.9) +
  scale_fill_viridis(option = "inferno", discrete = TRUE, begin = 0.1, end = 0.9) +
  labs(
    title = "Overview of HRV at Visit 1 by gender",
    x = "HRV measure", 
    y = "Z-score [SEM]"
  )
```

### HRV measures at V4

```{r}
# For males and females
tmp <- bind_rows(z_hrv4_male, z_hrv4_female)

# Create SEM
df <-
  cs4[c("ID", "HTN", "CHD", "DIABETES")] %>%
  inner_join(cs1[c("ID", "GENDER")], ., by = "ID") %>%
  gather(key = covar, value = status, -ID, -GENDER) %>%
  inner_join(gather(data = tmp[-3], key = hrv, value = value, -ID), ., by = "ID") %>%
  group_by(GENDER, covar, status, hrv) %>%
  na.omit() %>%
  summarise(mean = mean(value), n = length(value), sd = sd(value), se = sd(value) / sqrt(length(value)))

# Visualize
ggplot(df, aes(x=hrv, colour=hrv, shape = status)) + 
  facet_wrap(GENDER ~ covar) +  
  geom_point(aes(y = mean), size=3, position=position_dodge(0.3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2, position=position_dodge(0.3)) +
  theme_minimal() +
  scale_colour_viridis(option = "cividis", discrete = TRUE, begin = 0.1, end = 0.9) +
  scale_fill_viridis(option = "inferno", discrete = TRUE, begin = 0.1, end = 0.9) +
  labs(
    title = "Overview of HRV at Visit 4 by gender",
    x = "HRV measure", 
    y = "Z-score [SEM]"
  )
```
# CONCLUSION 

### References {.allowframebreaks}
