---
title: "Heart Rate Variability and Psychosocial States"
subtitle: "An Atherosclerotic Risk in Communities Study"
author: 
  - Anish Shah, MD^[Department of Medicine, School of Medicine, Emory University]
  - Amit Shah, MD, MSCR^[Department of Epidemiology, Rollins School of Public Health, Emory University]
data: "`r format(Sys.time(), '%B %d, %Y')`"
fontsize: 10pt
bibliography: hrv-aric.bib
csl: american-medical-association.csl
output: 
  beamer_presentation:
    colortheme: "beaver"
    slide_level: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, global_options, include=FALSE}

knitr::opts_chunk$set(
  cache = TRUE,
  warning = FALSE,
  eval = TRUE,
  echo = FALSE,
  include = TRUE,
  message = FALSE,
  dpi = 600,
  dev = "png",
  options("scipen" = 0, "digits" = 3),
  tinytex.verbose = TRUE
)

knitr::opts_knit$set(kable.force.latex=FALSE)
```

```{r}

# Libraries
source("1_libraries.R")

# Data intake
source("2_intake.R")

# Tidy
source("3_tidy.R")

```

# UNDERSTAND

## Introduction

### Disclosures and Funding

- Emory University
- Atherosclerosis Risk in Communities Study
- No conflicts of interest

### Atherosclerosis Risk in Communities Study

The ARIC study is a prospective epidemiologic study designed to investigate atherosclerosis and its clinical outcomes by cardiovascular risk factors, diseases, and demographics. It involved longitudinal follow-up over five visits from 1987 to 2013, with outcomes updated as of 2015, with over 15k initial participants. It collected measures of psychosocial stress (increased anger, increased vital exhaustion, decreased social support) and HRV over time.

### Background

Psychosocial stressors, such as fatigue and vital exhaustion, are repeatedly shown to be associated independently with MACE.[@Appels1988; @Bogle2018] Stressors and depression are shown to correlate with lower HRV.[@Huang2018; @Vroege2012; @Shah2013] The relationship between autonomic dysfunction and psychosocial stressors is still being studied.

**Does psychosocial stressor impact HRV longitudinally, and does their interaction reflect increased cardiovascular mortality?**

### Purpose

Objectives:

- Examine HRV from V1 and V4
- Study the relationship of anger from V2 to V4
- Cross-sectional relationship with vital exhaustion / social support at V2
- Compare these factors with longitudinal outcomes

Hypothesis:

- increased stress will associate with lower HRV
- changes in stress will associate with proportional changes in HRV
- the interaction of lower HRV and higher stress will have increased risk of mortality
  
## Methods

The data has been tidied up to this point. The current variables are:

Variable | Description
--- | ---
demo | Overall sample description, including death
cs2 | Prevalence of disease at V2
cs4 | Prevalence of disease at V4
hpa2 | ISEL and LSNS scores at V2
hpb2 | Maastricht vital exhaustion at V2
hpc2 | Spielberger Anger at V2
hpc4 | Spielberger Anger at V4
score_hp* | Numeric scaling of psychosocial scores
hrv1 | 2-minute HRV data from V1
hrv4 | 6-minute HRV data from V4

The general paradigm is that HRV will be modulated by stress.
**HRV ~ stress**

HRV remains the outcome variable
Emotional state is the exposure, which could be binary by median, or based on standard deviations. 

1. Exposed: high stress
1. Unexposed: low stress

# REPORT

## DESCRIPTION

The overall population should be described, but the at V2 and V4 they should be characterized by their exposure (the stressors). 

### Population

```{r, results = "asis"}

# THe initial population
char_v1 <- 
  compareGroups(~ BMI + HTN + CHD + DIABETES, data = cs1) %>%
  createTable(., type = 1, show.n = FALSE, hide.no = c("Absent"), digits = 0)

# Plot the table
export2md(char_v1)

```

### Interpersonal Support Evaluation List (ISEL)

The ISEL is scored from 0-3 points per question for 16-items, with the total ranging from 0-48 points (higher being more supportive). There are subscales, which are seen below as they map on to the HPAA questions.

Variable | Questions
--- | ---
Appraisal (AP) | 7, 10, 14, 17 
Tangible (TA) | 8, 9, 13, 16
Belonging (BE) | 4, 5, 6, 11, 18
Self-esteem (SE) | 3, 12, 15

The scores were normalized for ease of comparison.

### ISEL Distribution

```{r}

# Looking at scores from ISEL and subgroups
df <-
  score_hpa2[c("ID", "LSNS", "ISEL", "AP", "TA", "BE", "SE")] %>%
  gather(variable, value, -ID)

# ISEL normality
# Visualize the overall score distribution
# Mostly normal, slightly skewed
gDistributionInterpersonal <-
  ggplot(df[df$variable == "ISEL", ], aes(x = value)) +
  geom_histogram(binwidth = 1) +
  labs(
    title = "Distribution of ISEL total score",
    x = "Score",
    y = "Frequency"
  ) +
  theme_minimal()

# Visualize distribution of subscales
# Have variety of scews, do not "appear" normal
gDistributionInterpersonalSubscales <-
  ggplot(df[df$variable != "ISEL" & df$variable != "LSNS", ], aes(x = value, y = variable)) +
  geom_density_ridges2(aes(fill = variable), stat = "binline", binwidth = 1, colour = "white", scale = 1.6) +
  labs(
    title = "Distribution of ISEL subscales",
    x = "Points",
    y = "Frequency"
  ) +
  guides(fill = FALSE) +
  scale_fill_ptol() +
  theme_minimal()

# Plot the two ISEL graphs
grid.arrange(gDistributionInterpersonal, gDistributionInterpersonalSubscales, ncol = 2)

```

### Lubben Social Network Scale (LSNS)

The LSNS is scored from 0-5 points for an 10-item questionnaire (the original form), with the total ranging from 0-50 points (higher more supportive). The subscales haven't been validated. The scores were normalized for ease of comparison.

```{r}

# Looking at scores from LSNS
df <-
  score_hpa2[c("ID", "LSNS")] %>%
  gather(variable, value, -ID)

# Visualize the overall score distribution
# Mostly normal, slightly skewed
gDistributionLubben <-
  ggplot(df[df$variable == "LSNS", ], aes(x = value)) +
  geom_histogram(binwidth = 1) +
  labs(
    title = "Distribution of LSNS total score",
    x = "Score",
    y = "Frequency"
  ) +
  theme_minimal()

# Plot this
gDistributionLubben
```

### V2 Comparison by Social Support

```{r}

# Use the median for the HPA values for visit 2
# Split into quantiles based on the score (4 quantiles)
df <-
  within(score_hpa2, {
    iselgrp = ntile(ISEL, 4)
    apgrp = ntile(AP, 4)
    tagrp = ntile(TA, 4)
    begrp = ntile(BE, 4) 
    segrp = ntile(SE, 4)
    lsnsgrp = ntile(LSNS, 4)
  })

# New data table
df <- left_join(df, cs2, by = "ID")

# Comparison for ISEL
isel_v2 <-
  compareGroups(iselgrp ~ BMI + HTN + CHD + DIABETES + STROKE, data = df) %>%
  createTable(., type = 1, show.n = FALSE, digits = 1, hide.no = c("Absent"))

# Comparison for AP
ap_v2 <-
  compareGroups(apgrp ~ BMI + HTN + CHD + DIABETES + STROKE, data = df) %>%
  createTable(., type = 1, show.n = FALSE, digits = 1, hide.no = c("Absent"))

# Comparison for TA
ta_v2 <-
  compareGroups(tagrp ~ BMI + HTN + CHD + DIABETES + STROKE, data = df) %>%
  createTable(., type = 1, show.n = FALSE, digits = 1, hide.no = c("Absent"))

# Comparison for BE
be_v2 <-
  compareGroups(begrp ~ BMI + HTN + CHD + DIABETES + STROKE, data = df) %>%
  createTable(., type = 1, show.n = FALSE, digits = 1, hide.no = c("Absent"))

# Comparison for SE
se_v2 <-
  compareGroups(segrp ~ BMI + HTN + CHD + DIABETES + STROKE, data = df) %>%
  createTable(., type = 1, show.n = FALSE, digits = 1, hide.no = c("Absent"))

# Comparison for LSNS
lsns_v2 <-
  compareGroups(lsnsgrp ~ BMI + HTN + CHD + DIABETES + STROKE, data = df) %>%
  createTable(., type = 1, show.n = FALSE, digits = 1, hide.no = c("Absent"))

# Display below

```

`r export2md(isel_v2)`

----

`r export2md(ap_v2)`

----

`r export2md(ta_v2)`

----

`r export2md(be_v2)`

----

`r export2md(se_v2)`

----

`r export2md(lsns_v2)`

### Maastrich Vital Exhaustion Questionnaire

```{r}

# Overall Maastrich data frame
df <-
  score_hpb2 %>%
  gather(variable, value, -ID)

# Visualization
gDistributionMaastrich <-
  ggplot(df, aes(x = value)) +
  geom_histogram(aes(y = ..count.. / sum(..count..)), binwidth = 1) +
  labs(
    title = "Distribution of Maastrich Exhaustion Scale",
    x = "Score",
    y = "Frequency"
  ) +
  scale_y_continuous(labels = percent) +
  theme_minimal()
 
# Plot
gDistributionMaastrich
```

### V2 Comparison by Vital Exhaustion

```{r, results = "asis"}

# Maastrich scores broken down into 4 quantiles
# Higher the score, the more exhausted you are.
df <- 
  within(score_hpb2, {
    vitalgrp = ntile(MAASTRICH, 4)
  })

# New data table
df <- left_join(df, cs2, by = "ID")

# Comparison for ISEL
vital_v2 <-
  compareGroups(vitalgrp ~ BMI + HTN + CHD + DIABETES + STROKE, data = df) %>%
  createTable(., type = 1, show.n = FALSE, digits = 1, hide.no = c("Absent"))

# Data table
export2md(vital_v2)
```

### Spielberger Anger Trait Scale 

```{r}

# Overall Spielberger at V2
df <- score_hpc2

gDistributionSpielbergerV2 <-
  ggplot(df, aes(x = SPIELBERGER, fill = factor(anger_level))) +
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 1) +
  labs(
    title = "Distribution of Spielberger Anger Trait Scale [V2]",
    x = "Score",
    y = "Frequency",
    fill = "Anger Level"
  ) +
  scale_y_continuous(label = percent) +
  scale_fill_ptol() + 
  theme_minimal()

# Overall Spielberger at V2
df <- score_hpc4

gDistributionSpielbergerV4 <-
  ggplot(df, aes(x = SPIELBERGER, fill = factor(anger_level))) +
  geom_histogram(aes(y = ..count.. / sum(..count..)), binwidth = 1) +
  labs(
    title = "Distribution of Spielberger Anger Trait Scale [V4]",
    x = "Score",
    y = "Frequency",
    fill = "Anger Level"
  ) +
  scale_y_continuous(label = percent) +
  scale_fill_ptol() +
  theme_minimal()

# Plot side by side
grid.arrange(gDistributionSpielbergerV2, gDistributionSpielbergerV4, ncol = 1)
```

### V2 Comparison by Anger

```{r, results = "asis"}

# VISIT 2
# Spielberger anger trait broke down by anger level into quartiles
# Higher the score, the angrier you are
df <- 
  within(score_hpc2, {
    angergrp = ntile(SPIELBERGER, 4)
  })

# New data table
df <- left_join(df, cs2, by = "ID")

# Comparison for anger trait
anger_v2 <-
  compareGroups(angergrp ~ BMI + HTN + CHD + DIABETES + STROKE, data = df) %>%
  createTable(., type = 1, show.n = FALSE, digits = 1, hide.no = c("Absent"))

# Data table
export2md(anger_v2)

```

### V4 Comparison by Anger

```{r, results = "asis"}

# VISIT 4
# Spielberger anger trait broke down by anger level
# Higher the score, the angrier you are
df <- 
  within(score_hpc4, {
    angergrp = ntile(SPIELBERGER, 4)
  })

# New data table
df <- left_join(df, cs4, by = "ID")

# Comparison for anger trait
anger_v4 <-
  compareGroups(angergrp ~ BMI + HTN + CHD + DIABETES + STROKE, data = df) %>%
  createTable(., type = 1, show.n = FALSE, digits = 1, hide.no = c("Absent"))

# Display
export2md(anger_v4)
```

## HEART RATE VARIABILITY 

### V1 Distribution

```{r}

# Make a long format of variables for visualization
df <- hrv1 %>% gather(variable, value, -ID)

# Visualize each distribution pattern
gDistributionHRV1 <-
  ggplot(df, aes(x = value, fill = variable)) +
  facet_wrap(~ variable, ncol = 3, scales = "free") +
  geom_histogram(aes(y = ..count.. / sum(..count..))) +
  scale_y_continuous(labels = percent) +
  scale_fill_ptol() +
  labs(
    title = "Distribution of untransformed HRV [V1]",
    x = "Values",
    y = "Frequency (percent)"
  ) +
  guides(fill = FALSE) +
  theme_minimal()

# Normalized long table
df <- z_hrv1 %>% gather(variable, value, -ID)

# Ridge plots after normalization
gRidgesHRV1 <-
  ggplot(df, aes(x = value, y = variable)) +
  geom_density_ridges2(aes(fill = variable), colour = "white", scale = 1.3) +
  scale_fill_ptol() +
  xlim(-3, 3) +
  labs(
    title = "Distribution of HRV after normalization at V1",
    x = "Values after cube root and z-normalization",
    y = "HRV measures distribution"
  ) +
  guides(fill = FALSE) +
  theme_minimal()

# Correlation tests
df <- z_hrv1[-1]
correlation_plot_hrv_v1 <-
  ggcorr(df, nbreaks = 10, palette = "BrBG", label = TRUE)

# Plot
gRidgesHRV1
```

### V4 Distribution

```{r}

# Make a long format of variables for visualization
df <- hrv4 %>% gather(variable, value, -ID)

# Visualize each distribution pattern
gDistributionHRV4 <-
  ggplot(df, aes(x = value, fill = variable)) +
  facet_wrap(~ variable, ncol = 3, scales = "free") +
  geom_histogram(aes(y = ..count.. / sum(..count..))) +
  scale_y_continuous(labels = percent) +
  scale_fill_ptol() +
  labs(
    title = "Distribution of untransformed HRV [V4]",
    x = "Values",
    y = "Frequency (percent)"
  ) +
  guides(fill = FALSE) + 
  theme_minimal()

# Normalized long table
df <- z_hrv4 %>% gather(variable, value, -ID)

# Ridge plots after normalization
gRidgesHRV4 <-
  ggplot(df, aes(x = value, y = variable)) +
  geom_density_ridges2(aes(fill = variable), colour = "white", scale = 1.3) +
  scale_fill_ptol() +
  xlim(-3, 3) +
  labs(
    title = "Distribution of HRV after normalization at V4",
    x = "Values after cube root and z-normalization",
    y = "HRV measures distribution"
  ) +
  guides(fill = FALSE) +
  theme_minimal()

# Correlation tests
df <- z_hrv4[-1]
correlation_plot_hrv_v4 <-
  ggcorr(df, nbreaks = 10, palette = "BrBG", label = TRUE)

# Plot
gRidgesHRV4
```

## CROSS-SECTIONAL RELATIONSHIP of HRV and STRESS

### V2 Distribution

### V2 Linear Regressions

```{r}

# Data frame for V2
df <-
  merge(score_hpa2, score_hpb2, by = "ID") %>%
  merge(., score_hpc2, by = "ID") %>%
  gather(psych_variable, psych_value, -ID) %>%
  merge(., gather(z_hrv1, hrv_variable, hrv_value, -ID), by = "ID") %>%
  as_tibble()

# Create selection variables
df$psych_variable <- factor(df$psych_variable)
df$hrv_variable <- factor(df$hrv_variable)

# Create combination column as an identifier
df$combination <- paste(df$psych_variable, df$hrv_variable)

# Create combinatorial model
lm_v2 <-
  df %>%
  group_by(psych_variable, hrv_variable) %>%
  do(
    model =
      lm(.$psych_value ~ .$hrv_value, data = .)
  )

# Summarize the model and extract ocefficients
lm_v2$model_sum <- map(lm_v2$model, summary)
lm_v2$pvalue <-
  as.numeric(map(lm_v2$model_sum, function(x) {
    x$coefficients[2, 4]
  }))
lm_v2$rsquare <-
  as.numeric(map(lm_v2$model_sum, function(x) {
    x$r.squared
  }))

# Select only the significant models
lm_sig <- lm_v2[lm_v2$pvalue < .05, ]
lm_sig$combination <- paste(lm_sig$psych_variable, lm_sig$hrv_variable)

# Visualize regressions
gRegressionsV1 <-
  ggplot(df[df$combination %in% lm_sig$combination, ], aes(x = hrv_value, y = psych_value, colour = psych_variable)) +
  facet_wrap(~ hrv_variable, scales = "free", ncol = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  xlim(-3, 3) +
  labs(
    title = "Linear regressions of Psych ~ HRV at Visit 2",
    x = "Normalized HRV",
    y = "Normalized Psych Scores [SEM]",
    colour = "Psychosocial Measures",
    caption = "Significant correlations of HRV with psychosocial measures is visualized above."
  ) +
  scale_colour_ptol() +
  theme_minimal()

# Plot
gRegressionsV1
```


### V4 Distribution

### V4 Linear Regressions

```{r}

# Data from visit 4
df <-
  score_hpc4 %>%
  gather(psych_variable, psych_value, -ID) %>%
  merge(., gather(z_hrv4, hrv_variable, hrv_value, -ID)) %>%
  as_tibble()

# Create identifying variables
df$psych_variable <- factor(df$psych_variable)
df$hrv_variable <- factor(df$hrv_variable)
df$combination <- paste(df$psych_variable, df$hrv_variable)

# Create combinatorial model
lm_v4 <-
  df %>%
  group_by(psych_variable, hrv_variable) %>%
  do(
    model =
      lm(.$psych_value ~ .$hrv_value, data = .)
  )

# Summarize the model and extract ocefficients
lm_v4$model_sum <- map(lm_v4$model, summary)
lm_v4$pvalue <-
  as.numeric(map(lm_v4$model_sum, function(x) {
    x$coefficients[2, 4]
  }))
lm_v4$rsquare <-
  as.numeric(map(lm_v4$model_sum, function(x) {
    x$r.squared
  }))

# Select only the significant models
lm_sig <- lm_v4[lm_v4$pvalue < .05, ]
lm_sig$combination <- paste(lm_sig$psych_variable, lm_sig$hrv_variable)

# Visualize regressions
gRegressionsV4 <-
  ggplot(df[df$combination %in% lm_sig$combination, ], aes(x = hrv_value, y = psych_value, colour = psych_variable)) +
  facet_wrap(~ hrv_variable, scales = "free", ncol = 2) +
  geom_smooth(method = "lm", se = TRUE) +
  xlim(-3, 3) +
  labs(
    title = "Linear regressions of Psych ~ HRV at Visit 4",
    x = "Normalized HRV",
    y = "Normalized Psych Scores [SEM]",
    colour = "Psychosocial Measures"
  ) +
  scale_colour_ptol() +
  theme_minimal()

# Plot
gRegressionsV4
```




# CONCLUSION

### References {.allowframebreaks}



