---
title: "Heart Rate Variability and Psychosocial Stress"
subtitle: "An Atherosclerotic Risk in Communities Study"
author: "Anish Shah, MD"
data: "`r format(Sys.time(), '%B %d, %Y')`"
fontsize: 10pt
bibliography: hrv.bib
csl: american-medical-association.csl
link-citations: yes
output: 
  tufte::tufte_handout:
    latex_engine: xelatex
header-includes:
  - \usepackage{dcolumn}
  - \usepackage{float}
editor_options: 
  chunk_output_type: console
---

```{r setup, global_options, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
knitr::opts_chunk$set(
  cache = TRUE,
  warning = FALSE,
  eval = TRUE,
  echo = FALSE,
  include = TRUE,
  message = FALSE,
  dpi = 600,
  dev = "png",
  options("scipen" = 0, "digits" = 3),
  tinytex.verbose = TRUE,
  tidy = FALSE, 
  cache.extra = packageVersion('tufte')
)

options(xtable.comment = FALSE)
```

```{r, include=FALSE}

# Libraries
source("1_libraries.R")

# Data intake
source("2_intake.R")

# Tidy
source("3_tidy.R")
```

# OVERVIEW

## Disclosures and Funding

- Emory University
- Atherosclerosis Risk in Communities Study
- No conflicts of interest

## Atherosclerosis Risk in Communities Study

The ARIC study is a prospective epidemiologic study designed to investigate atherosclerosis and its clinical outcomes by cardiovascular risk factors, diseases, and demographics. It involved longitudinal follow-up over five visits from 1987 to 2013, with outcomes updated as of 2015, with over 15k initial participants. It collected measures of psychosocial stress (increased anger, increased vital exhaustion, decreased social support) and HRV over time.

## Background

Psychosocial stressors, such as fatigue and vital exhaustion, are repeatedly shown to be associated independently with MACE.[@Appels1988; @Bogle2018] Stressors and depression are shown to correlate with lower HRV.[@Huang2018; @Vroege2012; @Shah2013] The relationship between autonomic dysfunction and psychosocial stressors is still being studied. ^[What is the interaction between HRV and psychosocial stressors?]

## Purpose

Objectives:

- Examine HRV at V1 and V4
- Study cross-sectional relationship of stress and comorbidities
- Compare longitudinal pattern of anger from V2 and V4^[Will changes in anger associate with changes in HRV?]
- Study effect of stress on HRV^[Will increased stress associate with lower HRV?]
- Adjust for gender effect on stress^[Will gender and its effect on HRV influence the measurement of stress?]

## Methods

Statistical techniques^[Data description, student's t-test, linear regression, stepwise regression] will be varied to attempt to understand the frequentist and Bayesian influences of HRV, stress, and comorbidities. The data will be analyzed through R.^[Key packages include *tidyverse*, *ggplot2*, *MASS*, *lme4*.] 

# HRV description

## Vist 1 HRV measures

```{r, results='asis', include=FALSE}
# Data
df <- inner_join(hrv1, cs1, by = "ID") %>% na.omit()

# Corrected HRV
char_hrv_v1 <- tabular( (n=1) + Format(digits = 2)*(LF_HF + HFc + LFc + VLFc + TPc + SDNNc + RMSSDc)*(mean) ~ (GENDER + HTN + CHD + DIABETES), data = df )


# Comparison between comorbid conditions and corrected HRV
# GENDER
df %>%
  dplyr::select(ID, HFc, LFc, VLFc, TPc, LF_HF, SDNNc, RMSSDc, PNN50, GENDER) %>%
  gather(key = hrv, value = value, -ID, -GENDER) %>%
  group_by(GENDER, hrv) %>%
  summarise(value = list(value)) %>%
  spread(GENDER, value) %>%
  group_by(hrv) %>%
  mutate(p.value = t.test(unlist(Female), unlist(Male))$p.value,
         t.value = t.test(unlist(Female), unlist(Male))$statistic)

# HTN
df %>%
  dplyr::select(ID, HFc, LFc, VLFc, TPc, LF_HF, SDNNc, RMSSDc, HTN) %>%
  gather(key = hrv, value = value, -ID, -HTN) %>%
  group_by(HTN, hrv) %>%
  summarise(value = list(value)) %>%
  spread(HTN, value) %>%
  group_by(hrv) %>%
  mutate(p.value = t.test(unlist(`0`), unlist(`1`))$p.value,
         t.value = t.test(unlist(`0`), unlist(`1`))$statistic)

# CHD
df %>%
  dplyr::select(ID, HFc, LFc, VLFc, TPc, LF_HF, SDNNc, RMSSDc, CHD) %>%
  gather(key = hrv, value = value, -ID, -CHD) %>%
  group_by(CHD, hrv) %>%
  summarise(value = list(value)) %>%
  spread(CHD, value) %>%
  group_by(hrv) %>%
  mutate(p.value = t.test(unlist(`0`), unlist(`1`))$p.value,
         t.value = t.test(unlist(`0`), unlist(`1`))$statistic)

# DIABETES
df %>%
  dplyr::select(ID, HFc, LFc, VLFc, TPc, LF_HF, SDNNc, RMSSDc, DIABETES) %>%
  gather(key = hrv, value = value, -ID, -DIABETES) %>%
  group_by(DIABETES, hrv) %>%
  summarise(value = list(value)) %>%
  spread(DIABETES, value) %>%
  group_by(hrv) %>%
  mutate(p.value = t.test(unlist(`0`), unlist(`1`))$p.value,
         t.value = t.test(unlist(`0`), unlist(`1`))$statistic)

```


```{r}
# Table
char_hrv_v1 %>%
  as.matrix() %>%
  kable(., booktabs = TRUE)
```

```{r fig-margin, fig.margin=TRUE, fig.width=6.5, fig.height=4}
# Visualize differences
# Statistical differences seen at LF_HF **, LF *, SDNN *, VLF * (no difference in HF)
df %>%
  dplyr::select(ID, HFc, LFc, VLFc, TPc, LF_HF, SDNNc, RMSSDc, GENDER) %>%
  gather(key = hrv, value = value, -ID, -GENDER) %>%
  ggplot(data = ., aes(x = hrv, y = value, fill = GENDER)) +
  geom_boxplot(position = "dodge") + 
  scale_fill_viridis(option = "cividis", discrete = TRUE) +
  theme_minimal() +
  scale_y_log10() +
  annotate("text", x = 2, y = 1e-3, label = "**") + # LF_HF
  annotate("text", x = 3, y = 1e-3, label = "*") + # LF
  annotate("text", x = 5, y = 1e-3, label = "*") + # SDNN
  annotate("text", x = 7, y = 1e-3, label = "*") + # VLF (only measured for 2 minutes at V1)
  labs(
    title = "HRV by Gender at V1",
    x = "HRV measures",
    y = "Mean values"
  )
```

## Visit 4 HRV measures 

```{r, results='asis', include=FALSE}
# Data
df <- inner_join(hrv4, cs4, by = "ID") %>% inner_join(., cs1[c("GENDER", "ID")], by = "ID") %>% na.omit

# Corrected HRV
char_hrv_v4 <- tabular( (n=1) + Format(digits = 2)*(LF_HF + HFc + LFc + VLFc + TPc + SDNNc + RMSSDc)*(mean) ~ (GENDER + HTN + CHD + DIABETES), data = df )

# Comparison between comorbid conditions and corrected HRV
# GENDER
df %>%
  dplyr::select(ID, HFc, LFc, VLFc, TPc, LF_HF, SDNNc, RMSSDc, GENDER) %>%
  gather(key = hrv, value = value, -ID, -GENDER) %>%
  group_by(GENDER, hrv) %>%
  summarise(value = list(value)) %>%
  spread(GENDER, value) %>%
  group_by(hrv) %>%
  mutate(p.value = t.test(unlist(Female), unlist(Male))$p.value,
         t.value = t.test(unlist(Female), unlist(Male))$statistic)

# HTN
df %>%
  dplyr::select(ID, HFc, LFc, VLFc, TPc, LF_HF, SDNNc, RMSSDc, HTN) %>%
  gather(key = hrv, value = value, -ID, -HTN) %>%
  group_by(HTN, hrv) %>%
  summarise(value = list(value)) %>%
  spread(HTN, value) %>%
  group_by(hrv) %>%
  mutate(p.value = t.test(unlist(`0`), unlist(`1`))$p.value,
         t.value = t.test(unlist(`0`), unlist(`1`))$statistic)

# CHD
df %>%
  dplyr::select(ID, HFc, LFc, VLFc, TPc, LF_HF, SDNNc, RMSSDc, CHD) %>%
  gather(key = hrv, value = value, -ID, -CHD) %>%
  group_by(CHD, hrv) %>%
  summarise(value = list(value)) %>%
  spread(CHD, value) %>%
  group_by(hrv) %>%
  mutate(p.value = t.test(unlist(`0`), unlist(`1`))$p.value,
         t.value = t.test(unlist(`0`), unlist(`1`))$statistic)

# DIABETES
df %>%
  dplyr::select(ID, HFc, LFc, VLFc, TPc, LF_HF, SDNNc, RMSSDc, DIABETES) %>%
  gather(key = hrv, value = value, -ID, -DIABETES) %>%
  group_by(DIABETES, hrv) %>%
  summarise(value = list(value)) %>%
  spread(DIABETES, value) %>%
  group_by(hrv) %>%
  mutate(p.value = t.test(unlist(`0`), unlist(`1`))$p.value,
         t.value = t.test(unlist(`0`), unlist(`1`))$statistic)

```

The patient describes 

```{r}
# Table to support 
char_hrv_v4 %>%
  as.matrix() %>%
  kable(., booktabs = TRUE)
```

```{r, fig.margin=TRUE, fig.width=6.5, fig.height=4}

# Visualize differences
# Statistical differences seen at LF_HF **, LF *, SDNN **, VLF ** (no difference in HF)
df %>%
  dplyr::select(ID, HFc, LFc, VLFc, TPc, LF_HF, SDNNc, RMSSDc, GENDER) %>%
  gather(key = hrv, value = value, -ID, -GENDER) %>%
  ggplot(data = ., aes(x = hrv, y = value, fill = GENDER)) +
  geom_boxplot(position = "dodge") + 
  scale_fill_viridis(option = "cividis", discrete = TRUE) +
  theme_minimal() +
  scale_y_log10() +
  annotate("text", x = 2, y = 1e-3, label = "**") + # LF_HF
  annotate("text", x = 3, y = 1e-3, label = "*") + # LF
  annotate("text", x = 5, y = 1e-3, label = "**") + # SDNN
  annotate("text", x = 7, y = 1e-3, label = "**") + # VLF
  labs(
    title = "HRV by Gender at V4",
    x = "HRV measures",
    y = "Mean values"
  )
  
```

# Stress description

## Visit 2 Stress measures

```{r, results='asis', include = FALSE}
# Data
df <- inner_join(scores, cs2, by = "ID") %>% inner_join(., cs1[c("GENDER", "ID")], by = "ID") %>% na.omit

# Social support, exhaustion, and anger stressors
char_stress_v2 <- tabular( (n=1) + Format(digits = 2)*(ISEL +  AP + TA + BE + SE + LSNS + MAASTRICHT + SPIELBERGER)*(mean) ~ (GENDER + HTN + CHD + DIABETES), data = df )


# Compare stress markers, significant difference between gender 
# P value c- Bonferroni correction for interpretation
df %>% 
  dplyr::select(ID, ISEL, LSNS, AP, TA, BE, SE, MAASTRICHT, SPIELBERGER, GENDER) %>%
  gather(key = stress, value = value, -ID, -GENDER) %>%
  group_by(GENDER, stress) %>%
  summarise(value = list(value)) %>%
  spread(GENDER, value) %>%
  group_by(stress) %>%
  mutate(p.value = t.test(unlist(Female), unlist(Male))$p.value,
         t.value = t.test(unlist(Female), unlist(Male))$statistic)

# Compare all comorbid conditions 
# HTN
df %>% 
  dplyr::select(ID, ISEL, LSNS, AP, TA, BE, SE, MAASTRICHT, SPIELBERGER, HTN) %>%
  gather(key = stress, value = value, -ID, -HTN) %>%
  group_by(HTN, stress) %>%
  summarise(value = list(value)) %>%
  spread(HTN, value) %>%
  group_by(stress) %>%
  mutate(p.value = t.test(unlist(`0`), unlist(`1`))$p.value,
         t.value = t.test(unlist(`0`), unlist(`1`))$statistic)

# CHD
df %>% 
  dplyr::select(ID, ISEL, LSNS, AP, TA, BE, SE, MAASTRICHT, SPIELBERGER, CHD) %>%
  gather(key = stress, value = value, -ID, -CHD) %>%
  group_by(CHD, stress) %>%
  summarise(value = list(value)) %>%
  spread(CHD, value) %>%
  group_by(stress) %>%
  mutate(p.value = t.test(unlist(`0`), unlist(`1`))$p.value,
         t.value = t.test(unlist(`0`), unlist(`1`))$statistic)

# DIABETES
df %>% 
  dplyr::select(ID, ISEL, LSNS, AP, TA, BE, SE, MAASTRICHT, SPIELBERGER, DIABETES) %>%
  gather(key = stress, value = value, -ID, -DIABETES) %>%
  group_by(DIABETES, stress) %>%
  summarise(value = list(value)) %>%
  spread(DIABETES, value) %>%
  group_by(stress) %>%
  mutate(p.value = t.test(unlist(`0`), unlist(`1`))$p.value,
         t.value = t.test(unlist(`0`), unlist(`1`))$statistic)

# STROKE
df %>% 
  dplyr::select(ID, ISEL, LSNS, AP, TA, BE, SE, MAASTRICHT, SPIELBERGER, STROKE) %>%
  gather(key = stress, value = value, -ID, -STROKE) %>%
  group_by(STROKE, stress) %>%
  summarise(value = list(value)) %>%
  spread(STROKE, value) %>%
  group_by(stress) %>%
  mutate(p.value = t.test(unlist(`0`), unlist(`1`))$p.value,
         t.value = t.test(unlist(`0`), unlist(`1`))$statistic)

# Maastricht, TA, AP were consistently strongest differences b/w comorbid conditions.
```

```{r, results='asis'}
# Stress table
char_stress_v2 %>%
  as.matrix() %>%
  kable(., booktabs = TRUE)

# Visualize differences
# Maastricht **, Spielberger *, AP *, SE *, TA ** (no difference in social support or overall ISEL)
df %>%
  dplyr::select(ID, ISEL, LSNS, AP, TA, BE, SE, MAASTRICHT, SPIELBERGER, GENDER) %>%
  gather(key = stress, value = value, -ID, -GENDER) %>%
  ggplot(data = ., aes(x = stress, y = value, fill = GENDER)) +
  scale_y_log10() +
  geom_boxplot(position = "dodge") + 
  scale_fill_viridis(option = "cividis", discrete = TRUE) +
  theme_minimal() +
  annotate("text", x = 5, y = 5e-1, label = "**") + # Maastricht
  annotate("text", x = 7, y = 5e-1, label = "*") + # Spielberger
  annotate("text", x = 6, y = 5e-1, label = "*") + # SE
  annotate("text", x = 1, y = 5e-1, label = "*") + # AP
  annotate("text", x = 8, y = 5e-1, label = "**") + # TA
  labs(
    title = "Stress by Gender at V2",
    x = "Stress measures",
    y = "Mean values"
  )

```

## Visit 4 Stress measures

```{r, results='asis'}
# Data
df <- inner_join(score_hpc4, cs4, by = "ID") %>% inner_join(., cs1[c("GENDER", "ID")], by = "ID") %>% na.omit

# Anger stressors
char_stress_v4 <- tabular( (n=1) + Format(digits = 2)*(SPIELBERGER + anger_level)*(mean) ~ (GENDER + HTN + CHD + DIABETES), data = df )

char_stress_v4 %>%
  as.matrix() %>%
  kable(., booktabs = TRUE)
```

## Longitudinal anger comparison

```{r, results='asis'}
# Data
df <- inner_join(score_hpc4, score_hpc2, by = "ID") %>% na.omit
names(df) <- c("ID", "SPIELBERGER_v2", "anger_level_v2", "SPIELBERGER_v4", "anger_level_v4")

# Longitudinal impression of anger
char_stress_longitudinal <- tabular( (n=1) ~ Format(digits = 2) * (SPIELBERGER_v2 + anger_level_v2 + SPIELBERGER_v4 + anger_level_v4) * (mean), data = df)

# Comparison of groups overtime, paired t.test (statistically different).
tmp <- t.test(x = df$SPIELBERGER_v2, y = df$SPIELBERGER_v4, paired = TRUE, data = df)

# Visualization
ggplot(data = gather(df[c(1,2,4)], stress, value, -ID), aes(x = value, fill = stress)) + 
  geom_density(position = "identity") + 
  scale_fill_viridis(option="cividis", discrete=TRUE, alpha=.5) + 
  theme_minimal() +
  labs(
    title = "Anger scores at V2 and V4",
    x = "Spielberger anger score",
    y = "Frequency"
  ) +
  annotate("text", label = "Sig diff b/w means", x = 25, y = 0.1)
```


# RELATIONSHIPS

## HRV from V1 to Stress at V2

A potential effect will be if HRV changes occur before the psychosocial stressor is identified. Visit 1 had rudimentary ECG for 2-minutes, while Visit 2 was the documenation os psyschosocial states.^[Does HRV predict stress levels in the future?]

```{r, include = FALSE}
# Data in wide, to be used for all stressors
df <-
  hrv1 %>%
  inner_join(scores, ., by = "ID") %>%
  inner_join(cs1, ., by = "ID") %>%
  na.omit()

### ISEL

# Stepwise regression, using MASS package
full_model <- glm(ISEL ~ AGE_V1 + GENDER + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + HFc + LFc + VLFc + TPc + LF_HF + SDNNc + RMSSDc + PNN50, data = df)
step_model_ISEL <- stepAIC(full_model, direction = "both", trace = TRUE)
# Final model ... sig with smoking, LF, and Diabetes, and TP

### AP

# Stepwise regression, using MASS package
full_model <- glm(AP ~ AGE_V1 + GENDER + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + HFc + LFc + VLFc + TPc + LF_HF + SDNNc + RMSSDc + PNN50, data = df)
step_model_AP <- stepAIC(full_model, direction = "both", trace = TRUE)
# Final model... sig = BMI, drinking, smoking, HTN, HF, LF, TP
# Diabetes and race were also included


### TA

# Stepwise regression, using MASS package
full_model <- glm(TA ~ AGE_V1 + GENDER + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + HFc + LFc + VLFc + TPc + LF_HF + SDNNc + RMSSDc + PNN50, data = df)
step_model_TA <- stepAIC(full_model, direction = "both", trace = TRUE)
# Final model... sig = Age, Smoking, HTN, CHD, DM, LF_HF


### BE

# Stepwise regression, using MASS package
full_model <- glm(BE ~ AGE_V1 + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + HFc + LFc + VLFc + TPc + LF_HF + SDNNc + RMSSDc + PNN50, data = df)
step_model_BE <- stepAIC(full_model, direction = "both", trace = TRUE)
# Final model... sig = BMI, Smoking, HTN, DM, LF_HF

### SE

# Stepwise regression, using MASS package
full_model <- glm(SE ~ AGE_V1 + GENDER + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + HFc + LFc + VLFc + TPc + LF_HF + SDNNc + RMSSDc + PNN50, data = df)
step_model_SE <- stepAIC(full_model, direction = "both", trace = TRUE)
# Final model ... sig = race, drinking, HTN... no HRV measure mattered

### LSNS

# Stepwise regression, using MASS package
full_model <- glm(LSNS ~ AGE_V1 + GENDER + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + HFc + LFc + VLFc + TPc + LF_HF + SDNNc + RMSSDc + PNN50, data = df)
step_model_LSNS <- stepAIC(full_model, direction = "both", trace = TRUE)
# Final model ... sig = BMI, Smoking, DM, LF, TP, HTN

### MAASTRICHT

# Stepwise regression, using MASS package
full_model <- glm(MAASTRICHT ~ AGE_V1 + GENDER + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + HFc + LFc + VLFc + TPc + LF_HF + SDNNc + RMSSDc + PNN50, data = df)
step_model_MAASTRICHT <- stepAIC(full_model, direction = "both", trace = TRUE)

# Final model ... sig = BMI, DRINKER, smoking, HTN, CHD, DM, HF, LF_HF, SDNN, PNN50

### SPIELBERGER
# Stepwise regression, using MASS package
full_model <- glm(SPIELBERGER ~ AGE_V1 + GENDER + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + HFc + LFc + VLFc + TPc + LF_HF + SDNNc + RMSSDc + PNN50, data = df)
step_model_SPIELBERGER <- stepAIC(full_model, direction = "both", trace = TRUE)

# Final model ... sig = Age, BMI, race, drinker, smoker, CHD, HF, LF
```


```{r, results='asis'}
# Models are all collected after stepwise regression
stargazer(step_model_ISEL, 
          step_model_AP, 
          step_model_TA, 
          step_model_BE, 
          step_model_LSNS, 
          step_model_MAASTRICHT, 
          step_model_SPIELBERGER, 
          type = "latex",
          font.size = "scriptsize",
          column.sep.width = "0pt",
          header = FALSE,
          single.row = TRUE,
          align = TRUE,
          p.auto = FALSE,
          report = "vc*",
          digits = 2,
          model.numbers = FALSE,
          no.space = TRUE,
          omit = c("Constant", "RACEUnknown", "DRINKERUnknown", "SMOKERUnknown")
          )
```

## Stress at V2 to HRV at V4

Similarly, will psychosocial stressors lead to changes in HRV. The analysis was done in a similar fashion using a stepwise regression to find the optimum models.

```{r, include=FALSE}
# Data with stress and HRV
df <-
  hrv4 %>%
  inner_join(scores, ., by = "ID") %>%
  inner_join(cs2, ., by = "ID") %>%
  inner_join(cs1[c("ID", "GENDER", "RACE")], ., by = "ID") %>%
  na.omit()

# HFc
full_model <- glm(HFc ~ AGE_V2 + GENDER + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + STROKE + ISEL + AP + TA + BE + SE + LSNS + MAASTRICHT + SPIELBERGER, data = df)
step_model_HFc <- stepAIC(full_model, direction = "both", trace = TRUE)

# LFc
full_model <- glm(LFc ~ AGE_V2 + GENDER + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + STROKE + ISEL + AP + TA + BE + SE + LSNS + MAASTRICHT + SPIELBERGER, data = df)
step_model_LFc <- stepAIC(full_model, direction = "both", trace = TRUE)

# VLFc
full_model <- glm(VLFc ~ AGE_V2 + GENDER + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + STROKE + ISEL + AP + TA + BE + SE + LSNS + MAASTRICHT + SPIELBERGER, data = df)
step_model_VLFc <- stepAIC(full_model, direction = "both", trace = TRUE)

# TPc
full_model <- glm(TPc ~ AGE_V2 + GENDER + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + STROKE + ISEL + AP + TA + BE + SE + LSNS + MAASTRICHT + SPIELBERGER, data = df)
step_model_TPc <- stepAIC(full_model, direction = "both", trace = TRUE)

# LF_HF
full_model <- glm(LF_HF ~ AGE_V2 + GENDER + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + STROKE + ISEL + AP + TA + BE + SE + LSNS + MAASTRICHT + SPIELBERGER, data = df)
step_model_LF_HF <- stepAIC(full_model, direction = "both", trace = TRUE)

# SDNNc
full_model <- glm(SDNNc ~ AGE_V2 + GENDER + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + STROKE + ISEL + AP + TA + BE + SE + LSNS + MAASTRICHT + SPIELBERGER, data = df)
step_model_SDNNc <- stepAIC(full_model, direction = "both", trace = TRUE)

# RMSSDc
full_model <- glm(RMSSDc ~ AGE_V2 + GENDER + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + STROKE + ISEL + AP + TA + BE + SE + LSNS + MAASTRICHT + SPIELBERGER, data = df)
step_model_RMSSDc <- stepAIC(full_model, direction = "both", trace = TRUE)

# PNN50
full_model <- glm(PNN50 ~ AGE_V2 + GENDER + BMI + RACE + DRINKER + SMOKER + HTN + CHD + DIABETES + STROKE + ISEL + AP + TA + BE + SE + LSNS + MAASTRICHT + SPIELBERGER, data = df)
step_model_PNN50 <- stepAIC(full_model, direction = "both", trace = TRUE)

```

```{r, results='asis'}
# Models are all collected after stepwise regression
stargazer(step_model_HFc, 
          step_model_LFc, 
          step_model_VLFc, 
          step_model_TPc, 
          step_model_LF_HF, 
          step_model_SDNNc, 
          step_model_RMSSDc, 
          step_model_PNN50, 
          type = "latex",
          font.size = "scriptsize",
          column.sep.width = "0pt",
          header = FALSE,
          single.row = TRUE,
          align = TRUE,
          p.auto = FALSE,
          report = "vc*",
          digits = 2,
          model.numbers = FALSE,
          no.space = TRUE,
          omit = c("Constant", "RACEUnknown", "DRINKERUnknown", "SMOKERUnknown")
          )
```


# CONCLUSION 

## References
