---
title: "Heart Rate Variability and Psychosocial States"
subtitle: "An Atherosclerotic Risk in Communities Study"
author: 
- Anish Shah, MD^[Department of Medicine, School of Medicine, Emory University]
- Amit Shah, MD, MSCR^[Department of Epidemiology, Rollins School of Public Health, Emory University]
data: "`r format(Sys.time(), '%B %d, %Y')`"
fontsize: 10pt
bibliography: aric.bib
csl: american-medical-association.csl
output: 
  beamer_presentation:
    highlight: tango
    slide_level: 4
	colortheme: "dolphin"
editor_options: 
  chunk_output_type: console
---


# R workspace

## Project set-up

```{r setup, global_options, include=FALSE}

knitr::opts_chunk$set(
  cache = TRUE,
  warning = FALSE,
  eval = TRUE,
  echo = FALSE,
  include = TRUE,
  message = FALSE,
  results = "asis",
  dpi = 600,
  dev = "png",
  options("scipen" = 0, "digits" = 3)
)
```

## Versions

0.1.0 Wrangle - data intake first
0.2.0 Wrangle -  tidy the data
0.3.0 Model - data exploration
0.3.1 Model - relationships between variables
1.0.0 Report - initial preview of results
1.0.1 Report - reworking original document in a "presentation" mode

## Libraries

```{r}
start_time <- proc.time()

library(stargazer)
library(lubridate)
library(rio)
library(broom)
library(compareGroups)
library(icd)
library(reshape2)
library(ggridges)
library(scales)
library(ggthemes)
library(rcompanion)
library(GGally)
library(psych)
library(Hmisc)
library(RColorBrewer)
library(PerformanceAnalytics)
library(survival)
library(survminer)
library(magrittr)
library(tidyverse)
library(knitr)
library(kableExtra)
library(citr)
library(gridExtra)
```

# Wrangle

## Visit 1

```{r}

# Read in table
v1d <- read_csv("v1/DERIVE13.dat", col_names = TRUE)

# Select only important columns
svar <- c(
  "ID",
  "DRNKR01",
  "BMI01",
  "PRVCHD05",
  "MDDXMI02",
  "PREVMI05",
  "HXOFMI02",
  "DIABTS03",
  "HYPERT05",
  "HYPTMD01",
  "CHOLMDCODE01",
  "V1AGE01",
  "V1DATE01",
  "BIRTHDAT",
  "GENDER",
  "RACEGRP",
  "PLAQUE01",
  "CIGT01",
  "CIGTYR01"
)
v1d <- v1d[c(svar)]

# Format data
v1d$V1DATE01 <- dmy(v1d$V1DATE01)
v1d$BIRTHDAT <- dmy(v1d$BIRTHDAT)
```

## Visit 2

```{r}

# Selected variables
svar <- c(
  "ID",
  "DRNKR21",
  "BMI21",
  "PRVCHD21",
  "MDDXMI21",
  "HXOFMI21",
  "DIABTS23",
  "ECGMI24",
  "PRVSTR21",
  "HYPERT25",
  "HYPTMD21",
  "CHOLMDCODE21",
  "V2AGE22",
  "V2DATE21",
  "PLAQUE21",
  "CIGT21"
)

# Intake and format variables
v2d <- read_csv("v2/DERIVE2_10.dat", col_names = TRUE)
v2d <- v2d[c(svar)]
v2d$V2DATE21 <- dmy(v2d$V2DATE21)

## Intake of V2 psychosocial metrics (Health and Life Profile).

# Social support
hpa2 <- read_csv("v2/HPAA.dat", col_names = TRUE)
hpa2$HPAA30 <- dmy(hpa2$HPAA30)

# Vital exhaustion
hpb2 <- read_csv("v2/HPBA.dat", col_names = TRUE)
hpb2$HPBA22 <- dmy(hpb2$HPBA22)

# Spielberger anger trait
hpc2 <- read_csv("v2/HPCA.dat", col_names = TRUE)
hpc2$HPCA11 <- dmy(hpc2$HPCA11)
```

## Visit 4

```{r}

## Visit 4 data

# Selected variables.
svar <- c(
  "ID",
  "DRNKR41",
  "BMI41",
  "PRVCHD43",
  "MDDXMI41",
  "HXOFMI41",
  "DIABTS42",
  "ECGMI41",
  "PRVSTR41",
  "HYPERT45",
  "HYPTMD41",
  "CHOLMDCODE41",
  "V4AGE41",
  "V4DATE41",
  "PLAQUE42",
  "CIGT41"
)

v4d <- read_csv("v4/DERIVE46.dat", col_names = TRUE)
v4d <- v4d[c(svar)]
v4d$V4DATE41 <- dmy(v4d$V4DATE41)

## Psychosocial metric

# Spielberger anger trait at visit 4.
hpc4 <- read_csv("v4/HPCB04.dat", col_names = TRUE)
hpc4$HPCB11 <- dmy(hpc4$HPCB11)
```

## HRV data

HRV data was shared by Eric Whitsel from the ARIC group, containing HRV measures from visit 1 and visit 4. This includes both frequency-domain and time-domain methods that were derived by the ARIC group. No raw data was available.

```{r}

# Data from HRV from visit 1, supine data.

hrv1 <- read_csv("hrv/f_hrv11b.dat", col_names = TRUE)

# Variables
svar <- c(
  "ID",
  "POSTV11",
  "QCFLGV11",
  "M_HRV11",
  "HFV11",
  "LFV11",
  "VLFV11",
  "TPV11",
  "PNN50V11",
  "MSSDV11",
  "SDNNV11"
)
hrv1 <- hrv1[c(svar)]

## HRV data from visit 4.
hrv4 <- read_csv("hrv/hrv_re4f.dat", col_names = TRUE)

# Selected variables
svar <- c(
  "ID",
  "postV4r",
  "qcflgv4R",
  "m_hrv4r",
  "HFv4r",
  "LFv4r",
  "VLFv4r",
  "tpV4r",
  "mssdV4r",
  "pnn50V4r",
  "sdnnV4r"
)
hrv4 <- hrv4[c(svar)]
```

## Outcomes

Patient mortality and morbidity and time of death was available up to the year 2015. The outcomes were given alongside ICD codes.

```{r}

# Read in table
outcomes <- read_csv("outcomes/inc_by15.dat", col_names = TRUE)

# Select and trim data frame
svar <- c(
  "ID",
  "C7_DATEMI",
  "C7_DATEPROC",
  "UCOD",
  "DATED15",
  "DEAD15"
)
outcomes <- outcomes[c(svar)]

# Date of MI (12/31/2015 is just the censoring date)
outcomes$C7_DATEMI <- mdy(outcomes$C7_DATEMI)

# Date of catherization (12/31/2015 is a censoring date)
outcomes$C7_DATEPROC <- mdy(outcomes$C7_DATEPROC)

# Date of death
outcomes$DATED15 <- dmy(outcomes$DATED15)

# Clean up outcomes
outcomes <- outcomes[c("ID", "C7_DATEMI", "UCOD", "DATED15", "DEAD15")]
names(outcomes) <- c("ID", "DATE_MI", "UCOD", "DATE_DEAD", "DEAD")
```

# Tidy

## Demographics

```{r}

# Visit 1 prevalence
cs1 <- v1d

# BMI
cs1$BMI_1 <- round(cs1$BMI01, digits = 1)

# Diagnosis of hypertension, including treatment.
cs1$HTN_1 <- 0
cs1$HTN_1[v1d$HYPERT05 == 1 | v1d$HYPTMD01 == 1] <- 1

# Prior coronary artery disease by any known disease, MD decision, or ECG findings of Q waves.
cs1$CHD_1 <- 0
cs1$CHD_1[(v1d$PRVCHD05 | v1d$MDDXMI02 | v1d$HXOFMI02) == 1] <- 1

# Select variables
svar <- c("ID", "BIRTHDAT", "V1DATE01", "GENDER", "RACEGRP", "DRNKR01", "CIGT01", "BMI_1", "HTN_1", "CHD_1", "DIABTS03")
cs1 <- cs1[c(svar)]
names(cs1) <- c("ID", "DATE_BIRTH", "DATE_V1", "GENDER", "RACE", "DRINKER", "SMOKER", "BMI", "HTN", "CHD", "DIABETES")

cs1$AGE_V1 <- interval(cs1$DATE_V1, cs1$DATE_BIRTH) / years(1) 

# Convert to factors
cs1 <- within(cs1, {
  ID <- factor(ID)
 
  
  GENDER <- factor(GENDER)
  levels(GENDER) <- c("Female", "Male")

  RACE <- factor(RACE)
  levels(RACE) <- c("Asian", "Black", "Unknown", "White")

  DRINKER <- factor(DRINKER)
  levels(DRINKER) <- c("Current", "Former", "Never", "Unknown")

  SMOKER <- factor(SMOKER)
  levels(SMOKER) <- c("Current", "Former", "Never", "Unknown")

  CHD <- factor(CHD)
  levels(CHD) <- c("Absent", "Present")

  DIABETES <- factor(DIABETES)
  levels(DIABETES) <- c("Absent", "Present")

  HTN <- factor(HTN)
  levels(HTN) <- c("Absent", "Present")
})

# Clean
cs1 <- as_tibble(cs1)
```


## Visit 2 cross-sectional data

```{r}

# Will need to include BMI, drinking status, coronary disease, diabetes, CVA, hypertension, smoking status
cs2 <- v2d

# BMI
cs2$BMI_2 <- round(cs2$BMI21, digits = 1)
cs2$BMI_2cat <- cut(cs2$BMI21,
  breaks = c(-Inf, 18.5, 25, 30, 35, 40, Inf),
  labels = c("underweight", "normal", "overweight", "obese_mild", "obese_moderate", "obese_severe")
)

# Diagnosis of hypertension, including treatment.
cs2$HTN_2 <- 0
cs2$HTN_2[v2d$HYPERT25 == 1 | v2d$HYPTMD21 == 1] <- 1

# Prior coronary artery disease by any known disease, MD decision, or ECG findings of Q waves.
cs2$CHD_2 <- 0
cs2$CHD_2[(v2d$PRVCHD21 | v2d$MDDXMI21 | v2d$ECGMI24) == 1] <- 1

svar <- c("ID", "V2DATE21", "V2AGE22", "DRNKR21", "CIGT21", "BMI_2", "BMI_2cat", "HTN_2", "CHD_2", "DIABTS23", "PRVSTR21")
cs2 <- as_tibble(cs2[c(svar)])
names(cs2) <- c("ID", "DATE_V2", "AGE_V2", "DRINKER", "SMOKER", "BMI", "BMI_cat", "HTN", "CHD", "DIABETES", "STROKE")

# Convert to factors
cs2 <- within(cs2, {
  ID <- factor(ID)

  DRINKER <- factor(DRINKER)
  levels(DRINKER) <- c("Current", "Former", "Never", "Unknown")

  SMOKER <- factor(SMOKER)
  levels(SMOKER) <- c("Current", "Former", "Never", "Unknown")

  CHD <- factor(CHD)
  levels(CHD) <- c("Absent", "Present")

  DIABETES <- factor(DIABETES)
  levels(DIABETES) <- c("Absent", "Present")

  HTN <- factor(HTN)
  levels(HTN) <- c("Absent", "Present")

  STROKE <- factor(STROKE)
  levels(STROKE) <- c("Absent", "Present")
})

# Clean up
cs2 <- as_tibble(cs2)
```

## Visit 4 cross-sectional data

```{r}

# Make cross sectional data table
cs4 <- v4d

# Fix BMI
cs4$BMI_4 <- round(cs4$BMI41, digits = 1)
cs4$BMI_4cat <- cut(cs4$BMI41,
  breaks = c(-Inf, 18.5, 25, 30, 35, 40, Inf),
  labels = c("underweight", "normal", "overweight", "obese_mild", "obese_moderate", "obese_severe")
)

# Fix hypertension
cs4$HTN_4 <- 0
cs4$HTN_4[v4d$HYPERT45 == 1 | v4d$HYPTMD41 == 1] <- 1

# Prior coronary artery disease by any known disease, MD decision, or ECG findings of Q waves.
cs4$CHD_4 <- 0
cs4$CHD_4[(v4d$PRVCHD43 | v4d$MDDXMI41 | v4d$ECGMI41) == 1] <- 1

# Select variables
svar <- c("ID", "V4AGE41", "V4DATE41", "DRNKR41", "CIGT41", "CHD_4", "HTN_4", "BMI_4", "BMI_4cat", "DIABTS42", "PRVSTR41")
cs4 <- cs4[c(svar)]
names(cs4) <- c("ID", "AGE_V4", "DATE_V4", "DRINKER", "SMOKER", "CHD", "HTN", "BMI", "BMI_cat", "DIABETES", "STROKE")

# Convert to factors
cs4 <- within(cs4, {
  ID <- factor(ID)

  DRINKER <- factor(DRINKER)
  levels(DRINKER) <- c("Current", "Former", "Never", "Unknown")

  SMOKER <- factor(SMOKER)
  levels(SMOKER) <- c("Current", "Former", "Never", "Unknown")

  CHD <- factor(CHD)
  levels(CHD) <- c("Absent", "Present")

  DIABETES <- factor(DIABETES)
  levels(DIABETES) <- c("Absent", "Present")

  HTN <- factor(HTN)
  levels(HTN) <- c("Absent", "Present")

  STROKE <- factor(STROKE)
  levels(STROKE) <- c("Absent", "Present")
})

# Clean up
cs4 <- as_tibble(cs4)
```

## Longitudinal follow up
```{r}

# df need to have diagnoses organized, they are a mix of ICD-9 and ICD-10 codes. If preceded with a letter, is an ICD-10 code, and if starts with a number, is an ICD-9 code (mainly). Using "ICD" package for help.
df <- outcomes[c("ID", "UCOD")]
icd9 <- icd9cm_hierarchy
icd10 <- icd10cm2016

# Process as character strings
icd9[6:8] <- mapply(as.character, icd9[6:8])
icd10[6:8] <- mapply(as.character, icd10[6:8])

# icd-9 set up
codes9 <- df
codes9$icd9 <- NA
codes9$icd9[as.icd9(codes9$UCOD) %>% icd_is_valid()] <-
  codes9$UCOD[as.icd9(codes9$UCOD) %>% icd_is_valid()]
codes9$icd9 <- decimal_to_parts(codes9$icd9)$mjr

# Merge in codes
icd9$icd9 <- icd9$three_digit
codes9 <- unique(merge(codes9, icd9[c("icd9", "major", "sub_chapter", "chapter")], by = "icd9", all.x = TRUE))

# icd-10 set up
codes10 <- df
codes10$icd10 <- NA
codes10$icd10[as.icd10(codes10$UCOD) %>% icd_is_valid()] <-
  codes10$UCOD[as.icd10(codes10$UCOD) %>% icd_is_valid()]
codes10$icd10 <- decimal_to_parts(codes10$icd10)$mjr

# Merge in codes
icd10$icd10 <- icd10$three_digit
codes10 <- unique(merge(codes10, icd10[c("icd10", "major", "sub_chapter", "chapter")], by = "icd10", all.x = TRUE))

# Merge all the codes together
codes <- merge(codes9, codes10, by = "ID", all = TRUE)
codes$dx[!is.na(codes$icd9)] <- codes$major.x[!is.na(codes$icd9)]
codes$dx[!is.na(codes$icd10)] <- codes$major.y[!is.na(codes$icd10)]
codes$subchapter[!is.na(codes$icd9)] <- codes$sub_chapter.x[!is.na(codes$icd9)]
codes$subchapter[!is.na(codes$icd10)] <- codes$sub_chapter.y[!is.na(codes$icd10)]
codes$chapter[!is.na(codes$icd9)] <- codes$chapter.x[!is.na(codes$icd9)]
codes$chapter[!is.na(codes$icd10)] <- codes$chapter.y[!is.na(codes$icd10)]
codes <- codes %>% group_by(ID) %>% filter(row_number() == 1)
codes <- codes[c("ID", "icd9", "icd10", "dx", "subchapter", "chapter")]

# Add teh outcomes back in
df <- merge(outcomes, codes, by = "ID")

# Add age as well
df <- merge(df, cs1[c("ID", "DATE_BIRTH")], by = "ID")

# Add ages
df$AGE_DEATH <- interval(df$DATE_BIRTH, df$DATE_DEAD) / years(1)

# Final data
outcomes <- df[c("ID", "DATE_BIRTH", "DATE_MI", "DATE_DEAD", "DEAD", "AGE_DEATH", "UCOD", "icd9", "icd10", "dx", "subchapter", "chapter")] %>% as_tibble()
```

## Psychosocial measures

The HPAA file includes the Interpersonal Support and Evaluation List (ISEL) and the Lubben Social Network Scale (LSNS)

```{r}

# Select social support data
df <- hpa2

# ISEL is 3:18
# LSNS is 19:29

# ISEL variables
isel <- c("HPAA03", "HPAA04", "HPAA05", "HPAA06", "HPAA07", "HPAA08", "HPAA09", "HPAA10", "HPAA11", "HPAA12", "HPAA13", "HPAA14", "HPAA15", "HPAA16", "HPAA17", "HPAA18")

# LSNS variables
lsns <- c("HPAA19", "HPAA20", "HPAA21", "HPAA22", "HPAA23", "HPAA24", "HPAA25", "HPAA26", "HPAA27", "HPAA28", "HPAA29")

# Class of variables

# Trimmed HPAA data
df$placeholder <- NA
df <- df[c("ID", "placeholder", isel, lsns)]

# Create factors for ISEL
# Positive factors (higher score is better)
df[c(3, 6, 7, 8, 11, 12, 13, 14, 16, 18)] <-
  df[c(3, 6, 7, 8, 11, 12, 13, 14, 16, 18)] %>%
  mutate_all(funs(recode(., "A" = 0, "B" = 1, "C" = 2, "D" = 3)))

# Negative or reversed choices
df[c(4, 5, 9, 10, 15, 17)] <-
  df[c(4, 5, 9, 10, 15, 17)] %>%
  mutate_all(funs(recode(., "A" = 3, "B" = 2, "C" = 1, "D" = 0)))

# Create factors and scores for LSNS
df[c(19, 20, 21, 22, 23, 24)] <-
  df[c(19, 20, 21, 22, 23, 24)] %>%
  mutate_all(funs(recode(., "A" = 0, "B" = 1, "C" = 2, "D" = 3, "E" = 4, "F" = 5)))

df[c(25, 26)] <-
  df[c(25, 26)] %>%
  mutate_all(funs(recode(., "A" = 5, "B" = 4, "C" = 3, "D" = 2, "E" = 1, "F" = 0)))

df <- within(df, {
  HPAA27 <- recode(HPAA27, "A" = 1, "B" = 0)
  HPAA28 <- recode(HPAA28, "B" = 5, "C" = 4, "D" = 3, "E" = 2, "F" = 1)
  HPAA29 <- recode(HPAA29, "A" = 4, "B" = 3, "C" = 2, "D" = 1)
})

# Summarize data based on scale
df$ISEL <- rowSums(df[c(3:18)], na.rm = TRUE)
df$LSNS <- rowSums(df[c(19:29)], na.rm = TRUE)

# Summarize data with ISEL subscales
df$AP <- rowSums(df[c(7, 10, 14, 17)], na.rm = TRUE)
df$TA <- rowSums(df[c(8, 9, 13, 16)], na.rm = TRUE)
df$BE <- rowSums(df[c(4, 5, 6, 11, 18)], na.rm = TRUE)
df$SE <- rowSums(df[c(3, 12, 15)], na.rm = TRUE)

# Place data back
score_hpa2 <- df[c("ID", "ISEL", "LSNS", "AP", "TA", "BE", "SE")]

# Normalize the data
score_hpa2[-1] <-
  as_tibble(scale(score_hpa2[-1], center = TRUE), scale = TRUE)
```

The HPBA file includes the Maastricht Vital Exhaustion Questionnaire.

```{r}

# Maastrich data
# Don't know is considered NA
df <- hpb2

# HPBA variable set up (questions number align)
questions <- c("HPBA01", "HPBA02", "HPBA03", "HPBA04", "HPBA05", "HPBA06", "HPBA07", "HPBA08", "HPBA09", "HPBA10", "HPBA11", "HPBA12", "HPBA13", "HPBA14", "HPBA15", "HPBA16", "HPBA17", "HPBA18", "HPBA19", "HPBA20", "HPBA21")
df <- df[c(questions, "ID")]

# Mutate to scores
df[c(1:8, 10:13, 15:21)] <-
  df[c(1:8, 10:13, 15:21)] %>%
  mutate_all(funs(recode(., "Y" = 1, "N" = 0)))

# Reversed scores
df[c(9, 14)] <-
  df[c(9, 14)] %>%
  mutate_all(funs(recode(., "Y" = 0, "N" = 1)))

df$MAASTRICH <- rowSums(df[c(1:21)], na.rm = TRUE)

# Place data back
score_hpb2 <- df[c("ID", "MAASTRICH")]
score_hpb2[-1] <-
  as_tibble(scale(score_hpb2[-1], center = TRUE, scale = TRUE))
```

This is the HPCB file from both visit 2 and visit 4. This includes the Spielberger anger trait scale, which ranged in scores from 10 to 40. 

```{r}

# Visit 2 HPCB file 
df <- hpc2

# Create data frame from Spielberger, trimmed down
questions <- c("HPCA01", "HPCA02", "HPCA03", "HPCA04", "HPCA05", "HPCA06", "HPCA07", "HPCA08", "HPCA09", "HPCA10")
df <- df[c(questions, "ID")]

# Mutate scores, with low = 1, high = 4
df[c(1:10)] <-
  df[c(1:10)] %>%
  mutate_all(funs(recode(., "A" = 1, "B" = 2, "C" = 3, "D" = 4)))

# Create total score
df$SPIELBERGER <- rowSums(df[c(1:10)], na.rm = TRUE)

# Create strata defined as high = [22, 40], mid = [15, 21], low = [10,14]
# Variable "anger_level" shows the three strata (low = 1, mid = 2, high = 3)

df$anger_level <- 1
df$anger_level[df$SPIELBERGER >= 15] <- 2
df$anger_level[df$SPIELBERGER >= 22] <- 3

# Confirm data frame, scaling for normality
score_hpc2 <- df[c("ID", "SPIELBERGER", "anger_level")]
score_hpc2[-c(1, 3)] <-
  as_tibble(scale(score_hpc2[-c(1, 3)], center = TRUE, scale = TRUE))

# Visit 4 HPCB file
df <- hpc4

# Create data frame from Maastrich, trimmed down
questions <- c("HPCB1", "HPCB2", "HPCB3", "HPCB4", "HPCB5", "HPCB6", "HPCB7", "HPCB8", "HPCB9", "HPCB10")
df <- df[c(questions, "ID")]

# Mutate scores, with low = 1, high = 4
df[c(1:10)] <-
  df[c(1:10)] %>%
  mutate_all(funs(recode(., "A" = 1, "B" = 2, "C" = 3, "D" = 4)))

# Create total score
df$SPIELBERGER <- rowSums(df[c(1:10)], na.rm = TRUE)

# Create strata defined as high = [22, 40], mid = [15, 21], low = [10,14]
# Variable "anger_level" shows the three strata (low = 1, mid = 2, high = 3)

df$anger_level <- 1
df$anger_level[df$SPIELBERGER >= 15] <- 2
df$anger_level[df$SPIELBERGER >= 22] <- 3

# Confirm data frame
score_hpc4 <- df[c("ID", "SPIELBERGER", "anger_level")]
score_hpc4[-c(1, 3)] <-
  as_tibble(scale(score_hpc4[-c(1, 3)], center = TRUE, scale = TRUE))
```

## Heart rate variability

During visit 1, HRV was taken in both supine and standing. The standard is to use the supine data. This was collected over 2 minutes only. The data here is normalized using a cube root method to avoid "infinite" values during scaling (which would occur using log-transformation). They were then scaled using the traditional z-transformation centered around 0.

```{r}

# Visit 1 HRV data
df <- hrv1

# Variables and data trim
svar <- c(
  "ID",
  "M_HRV11",
  "HFV11",
  "LFV11",
  "VLFV11",
  "TPV11",
  "PNN50V11",
  "MSSDV11",
  "SDNNV11"
)
df <- df[c("QCFLGV11", svar)]

# Only "quality" data
df <- df[df$QCFLGV11 != 1, svar]

# Rename the columns for ease of use
names(df) <- c("ID", "BPM", "HF", "LF", "VLF", "TP", "PNN50", "RMSSD", "SDNN")

# Only quality HRV data
hrv1 <- df

# Will need to transform the data prior to utilization
# Cube root to avoid the zero issues and negatives for scaling
z_hrv1 <-
  hrv1 %>%
  mutate_at(
    function(x) {
      sign(x) * abs(x)^(1 / 3)
      (x - mean(x)) / sd(x)
    },
    .vars = vars(-ID)
  )
```

At visit 4, HRV was collected again while supine. The recordings are 6 minutes in length. The values were transformed by the cube root before scaling using z-transformation centered around 0.

```{r}

# Select data frame
df <- hrv4

# Variable selection
svar <- c(
  "ID",
  "m_hrv4r",
  "HFv4r",
  "LFv4r",
  "VLFv4r",
  "tpV4r",
  "pnn50V4r",
  "mssdV4r",
  "sdnnV4r"
)
df <- df[c("qcflgv4R", svar)]

# Only quality data
df <- df[df$qcflgv4R != 1, svar]

# Rename variables
names(df) <- c("ID", "BPM", "HF", "LF", "VLF", "TP", "PNN50", "RMSSD", "SDNN")

# Final data frame
hrv4 <- df

# Will need to transform the data prior to utilization

# Cube root to avoid the zero issues and negatives for scaling
z_hrv4 <-
  hrv4 %>%
  na.omit() %>%
  mutate_at(
    function(x) {
      sign(x) * abs(x)^(1 / 3)
      (x - mean(x)) / sd(x)
    },
    .vars = vars(-ID)
  )
```

## Clean up

```{r}

# Clean up of memory-clogging junk
rm(df, questions, svar, v1d, v2d, v4d)
```

# Understand

## Introduction

### Disclosures and Funding

- Emory University
- Atherosclerosis Risk in Communities Study
- No conflicts of interest

### Atherosclerosis Risk in Communities Study

The ARIC study is a prospective epidemiologic study designed to investigate atherosclerosis and its clinical outcomes by cardiovascular risk factors, diseases, and demographics. It involved longitudinal follow-up over five visits from 1987 to 2013, with outcomes updated as of 2015, with over 15k initial participants. It collected measures of psychosocial stress (increased anger, increased vital exhaustion, decreased social support) and HRV over time.

### Background

Psychosocial stressors, such as fatigue and vital exhaustion, are repeatedly shown to be associated independently with MACE.[@Appels1988; @Bogle2018] Stressors and depression are shown to correlate with lower HRV.[@Huang2018; @Vroege2012; @Shah2013] The relationship between autonomic dysfunction and psychosocial stressors is still being studied.

**Does psychosocial stressor impact HRV longitudinally, and does their interaction reflect increased cardiovascular mortality?**

### Purpose

Objectives:

- Examine HRV from V1 and V4
- Study the relationship of anger from V2 to V4
- Cross-sectional relationship with vital exhaustion / social support at V2
- Compare these factors with longitudinal outcomes

Hypothesis:

- increased stress will associate with lower HRV
- changes in stress will associate with proportional changes in HRV
- the interaction of lower HRV and higher stress will have increased risk of mortality
  
## Methods

The data has been tidied up to this point. The current variables are...

Variable | Description
--- | ---
demo | Overall sample description, including death
cs2 | Prevalence of disease at V2
cs4 | Prevalence of disease at V4
hpa2 | ISEL and LSNS scores at V2
hpb2 | Maastricht vital exhaustion at V2
hpc2 | Spielberger Anger at V2
hpc4 | Spielberger Anger at V4
score_hp* | Numeric scaling of psychosocial scores
hrv1 | 2-minute HRV data from V1
hrv4 | 6-minute HRV data from V4

# Report

## Description

The overall population should be described, but the at V2 and V4 they should be characterized by their exposure (the stressors). 


### Population

```{r}

# THe initial population
char_v1 <- 
  compareGroups(~ GENDER + RACE + BMI + HTN + CHD + DIABETES, data = cs1) %>%
  createTable(., type = 1, show.n = FALSE, hide.no = c("Absent"), digits = 0)

# Plot the table
export2md(char_v1)

```


### Psychosocial measures

### Interpersonal support evaluation list (ISEL)

The ISEL is scored from 0-3 points per question for 16-items, with the total ranging from 0-48 points (higher being more supportive). There are subscales, which are seen below as they map on to the HPAA questions.

Variable | Questions
--- | ---
Appraisal (AP) | 7, 10, 14, 17 
Tangible (TA) | 8, 9, 13, 16
Belonging (BE) | 4, 5, 6, 11, 18
Self-esteem (SE) | 3, 12, 15

The scores were normalized for ease of comparison.

### ISEL distribution

```{r}

# Looking at scores from ISEL and subgroups
df <-
  score_hpa2[c("ID", "LSNS", "ISEL", "AP", "TA", "BE", "SE")] %>%
  gather(variable, value, -ID)

# ISEL normality
# Visualize the overall score distribution
# Mostly normal, slightly skewed
gDistributionInterpersonal <-
  ggplot(df[df$variable == "ISEL", ], aes(x = value)) +
  geom_histogram(binwidth = 1) +
  labs(
    title = "Distribution of ISEL total score",
    x = "Score",
    y = "Frequency"
  ) +
  theme_minimal()

# Visualize distribution of subscales
# Have variety of scews, do not "appear" normal
gDistributionInterpersonalSubscales <-
  ggplot(df[df$variable != "ISEL" & df$variable != "LSNS", ], aes(x = value, y = variable)) +
  geom_density_ridges2(aes(fill = variable), stat = "binline", binwidth = 1, colour = "white", scale = 1.6) +
  labs(
    title = "Distribution of ISEL subscales",
    x = "Points",
    y = "Frequency"
  ) +
  guides(fill = FALSE) +
  scale_fill_ptol() +
  theme_minimal()

# Plot the two ISEL graphs
grid.arrange(gDistributionInterpersonal, gDistributionInterpersonalSubscales, ncol = 2)

```

### Lubben social network scale (LSNS)

The LSNS is scored from 0-5 points for an 10-item questionnaire (the original form), with the total ranging from 0-50 points (higher more supportive). The subscales haven't been validated. The scores were normalized for ease of comparison.

```{r}

# Looking at scores from LSNS
df <-
  score_hpa2[c("ID", "LSNS")] %>%
  gather(variable, value, -ID)

# Visualize the overall score distribution
# Mostly normal, slightly skewed
gDistributionLubben <-
  ggplot(df[df$variable == "LSNS", ], aes(x = value)) +
  geom_histogram(binwidth = 1) +
  labs(
    title = "Distribution of LSNS total score",
    x = "Score",
    y = "Frequency"
  ) +
  theme_minimal()

# Plot this
gDistributionLubben
```

### Maastrich vital exhaustion questionnaire

```{r}

# Overall Maastrich data frame
df <-
  score_hpb2 %>%
  gather(variable, value, -ID)

# Visualization
gDistributionMaastrich <-
  ggplot(df, aes(x = value)) +
  geom_histogram(aes(y = ..count.. / sum(..count..)), binwidth = 1) +
  labs(
    title = "Distribution of Maastrich Exhaustion Scale",
    x = "Score",
    y = "Frequency"
  ) +
  scale_y_continuous(labels = percent) +
  theme_minimal()
 
# Plot
gDistributionMaastrich
```


### Spielberger anger trait scale

```{r}

# Overall Spielberger at V2
df <- score_hpc2

gDistributionSpielbergerV2 <-
  ggplot(df, aes(x = SPIELBERGER, fill = factor(anger_level))) +
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 0.5) +
  labs(
    title = "Distribution of Spielberger Anger Trait Scale [V2]",
    x = "Score",
    y = "Frequency",
    fill = "Anger Level"
  ) +
  scale_y_continuous(label = percent) +
  scale_fill_ptol() + 
  theme_minimal()

# Overall Spielberger at V2
df <- score_hpc4

gDistributionSpielbergerV4 <-
  ggplot(df, aes(x = SPIELBERGER, fill = factor(anger_level))) +
  geom_histogram(aes(y = ..count.. / sum(..count..)), binwidth = 0.5) +
  labs(
    title = "Distribution of Spielberger Anger Trait Scale [V4]",
    x = "Score",
    y = "Frequency",
    fill = "Anger Level"
  ) +
  scale_y_continuous(label = percent) +
  scale_fill_ptol() +
  theme_minimal()

# Plot side by side
grid.arrange(gDistributionSpielbergerV2, gDistributionSpielbergerV4, ncol = 1)
```

## Heart rate variability measures


### Distribution at visit 1

```{r}

# Make a long format of variables for visualization
df <- hrv1 %>% gather(variable, value, -ID)

# Visualize each distribution pattern
gDistributionHRV1 <-
  ggplot(df, aes(x = value, fill = variable)) +
  facet_wrap(~ variable, ncol = 3, scales = "free") +
  geom_histogram(aes(y = ..count.. / sum(..count..))) +
  scale_y_continuous(labels = percent) +
  scale_fill_ptol() +
  labs(
    title = "Distribution of untransformed HRV [V1]",
    x = "Values",
    y = "Frequency (percent)"
  ) +
  guides(fill = FALSE) +
  theme_minimal()

# Normalized long table
df <- z_hrv1 %>% gather(variable, value, -ID)

# Ridge plots after normalization
gRidgesHRV1 <-
  ggplot(df, aes(x = value, y = variable)) +
  geom_density_ridges2(aes(fill = variable), colour = "white", scale = 1.3) +
  scale_fill_ptol() +
  xlim(-3, 3) +
  labs(
    title = "Distribution of HRV after normalization at V1",
    x = "Values after cube root and z-normalization",
    y = "HRV measures distribution"
  ) +
  guides(fill = FALSE) +
  theme_minimal()

# Correlation tests
df <- z_hrv1[-1]
correlation_plot_hrv_v1 <-
  ggcorr(df, nbreaks = 10, palette = "BrBG", label = TRUE)

# Plot
gRidgesHRV1
```

### Distribution at visit 4

```{r}

# Make a long format of variables for visualization
df <- hrv4 %>% gather(variable, value, -ID)

# Visualize each distribution pattern
gDistributionHRV4 <-
  ggplot(df, aes(x = value, fill = variable)) +
  facet_wrap(~ variable, ncol = 3, scales = "free") +
  geom_histogram(aes(y = ..count.. / sum(..count..))) +
  scale_y_continuous(labels = percent) +
  scale_fill_ptol() +
  labs(
    title = "Distribution of untransformed HRV [V4]",
    x = "Values",
    y = "Frequency (percent)"
  ) +
  guides(fill = FALSE) + 
  theme_minimal()

# Normalized long table
df <- z_hrv4 %>% gather(variable, value, -ID)

# Ridge plots after normalization
gRidgesHRV4 <-
  ggplot(df, aes(x = value, y = variable)) +
  geom_density_ridges2(aes(fill = variable), colour = "white", scale = 1.3) +
  scale_fill_ptol() +
  xlim(-3, 3) +
  labs(
    title = "Distribution of HRV after normalization at V4",
    x = "Values after cube root and z-normalization",
    y = "HRV measures distribution"
  ) +
  guides(fill = FALSE) +
  theme_minimal()

# Correlation tests
df <- z_hrv4[-1]
correlation_plot_hrv_v4 <-
  ggcorr(df, nbreaks = 10, palette = "BrBG", label = TRUE)

# Plot
gRidgesHRV4
```



## Relationship between stress and HRV


### Visit 2 relationships

```{r}

# Data frame for V2
df <-
  merge(score_hpa2, score_hpb2, by = "ID") %>%
  merge(., score_hpc2, by = "ID") %>%
  gather(psych_variable, psych_value, -ID) %>%
  merge(., gather(z_hrv1, hrv_variable, hrv_value, -ID), by = "ID") %>%
  as_tibble()

# Create selection variables
df$psych_variable <- factor(df$psych_variable)
df$hrv_variable <- factor(df$hrv_variable)

# Create combination column as an identifier
df$combination <- paste(df$psych_variable, df$hrv_variable)

# Create combinatorial model
lm_v2 <-
  df %>%
  group_by(psych_variable, hrv_variable) %>%
  do(
    model =
      lm(.$psych_value ~ .$hrv_value, data = .)
  )

# Summarize the model and extract ocefficients
lm_v2$model_sum <- map(lm_v2$model, summary)
lm_v2$pvalue <-
  as.numeric(map(lm_v2$model_sum, function(x) {
    x$coefficients[2, 4]
  }))
lm_v2$rsquare <-
  as.numeric(map(lm_v2$model_sum, function(x) {
    x$r.squared
  }))

# Select only the significant models
lm_sig <- lm_v2[lm_v2$pvalue < .05, ]
lm_sig$combination <- paste(lm_sig$psych_variable, lm_sig$hrv_variable)

# Visualize regressions
gRegressionsV1 <-
  ggplot(df[df$combination %in% lm_sig$combination, ], aes(x = hrv_value, y = psych_value, colour = psych_variable)) +
  facet_wrap(~ hrv_variable, scales = "free", ncol = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  xlim(-3, 3) +
  labs(
    title = "Linear regressions of Psych ~ HRV at Visit 2",
    x = "Normalized HRV",
    y = "Normalized Psych Scores [SEM]",
    colour = "Psychosocial Measures",
    caption = "Significant correlations of HRV with psychosocial measures is visualized above."
  ) +
  scale_colour_ptol() +
  theme_minimal()

# Plot
gRegressionsV1
```

### Visit 4 Relationship

```{r}

# Data from visit 4
df <-
  score_hpc4 %>%
  gather(psych_variable, psych_value, -ID) %>%
  merge(., gather(z_hrv4, hrv_variable, hrv_value, -ID)) %>%
  as_tibble()

# Create identifying variables
df$psych_variable <- factor(df$psych_variable)
df$hrv_variable <- factor(df$hrv_variable)
df$combination <- paste(df$psych_variable, df$hrv_variable)

# Create combinatorial model
lm_v4 <-
  df %>%
  group_by(psych_variable, hrv_variable) %>%
  do(
    model =
      lm(.$psych_value ~ .$hrv_value, data = .)
  )

# Summarize the model and extract ocefficients
lm_v4$model_sum <- map(lm_v4$model, summary)
lm_v4$pvalue <-
  as.numeric(map(lm_v4$model_sum, function(x) {
    x$coefficients[2, 4]
  }))
lm_v4$rsquare <-
  as.numeric(map(lm_v4$model_sum, function(x) {
    x$r.squared
  }))

# Select only the significant models
lm_sig <- lm_v4[lm_v4$pvalue < .05, ]
lm_sig$combination <- paste(lm_sig$psych_variable, lm_sig$hrv_variable)

# Visualize regressions
gRegressionsV4 <-
  ggplot(df[df$combination %in% lm_sig$combination, ], aes(x = hrv_value, y = psych_value, colour = psych_variable)) +
  facet_wrap(~ hrv_variable, scales = "free", ncol = 2) +
  geom_smooth(method = "lm", se = TRUE) +
  xlim(-3, 3) +
  labs(
    title = "Linear regressions of Psych ~ HRV at Visit 4",
    x = "Normalized HRV",
    y = "Normalized Psych Scores [SEM]",
    colour = "Psychosocial Measures"
  ) +
  scale_colour_ptol() +
  theme_minimal()

# Plot
gRegressionsV4
```




# Conclusion

## Conclusion
### Bibliography

```{r}

# Time needed to process entire script
proc.time() - start_time
```



